---
title: "ATUS Clustering"
date: "`r format(Sys.time(), '%B %d %Y')`"
output:
  pdf_document:
    toc: true
    toc_depth: 2
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(message = FALSE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_knit$set(root.dir = normalizePath(".."))
```

```{r include=FALSE}
get_cluster_polygons <- function(df, x, y, cluster){
  # function returns polygons to draw around plot
  # usage: ggplot() + geom_polygon(data = hulls) +
  # x, y, and cluster must be strings
  
  find_hull <- function(df) df[chull(df[[x]], df[[y]]), ]
  hulls <- plyr::ddply(df, cluster, find_hull)
  
  return(hulls)
}
```


```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(pander)
library(dendextend)
library(NbClust)
options(mc.cores = parallel::detectCores())
set.seed(44)
theme_set(theme_minimal())

atus_long <- read_tsv('Data/atus.tsv')
demographics <- read_tsv('Data/demographic.tsv')
demographics <- demographics %>% 
  mutate(has_child = as.numeric(n_child > 0)) %>% 
  select(-c('age_youngest', 'n_child', 'state'))
```


# EDA

Note: Only contains weekend observations

## The data and its distribution
Note: redo this with the sample data??

```{r}
# description of the data
atus_long %>% 
  group_by(description) %>% 
  summarize(Mean = mean(value),
            part_rate = mean(value > 0)) %>% 
  arrange(desc(Mean)) %>% 
  mutate(Type = 'Continuous') %>% 
  janitor::adorn_totals() %>% 
  mutate(Mean = round(Mean, 0),
         part_rate = round(part_rate, 2),
         part_rate = ifelse(part_rate == 6.34, '-', part_rate)) %>% 
  select(Activity = description, Type, Mean, 'Participation rate' = part_rate) %>% 
  pander::pander(justify = c("left", "left", 'right', 'right'))
```

```{r}
# density plot of activities
atus_long %>% 
  group_by(description) %>% 
  mutate(Mean = mean(value)) %>% 
  ungroup() %>% 
  mutate(description = paste0(round(Mean, 1), ' min: ', description)) %>% 
  ggplot(aes(x = value, group = reorder(description, -Mean), fill = reorder(description, -Mean))) +
  geom_density(alpha = 0.4) +
  scale_fill_discrete(name = 'Mean minutes: Activity') +
  scale_x_log10() +
  labs(title = 'Time use activities',
       # subtitle = 'Non-zero values only',
       x = NULL,
       y = NULL) +
  theme(legend.position = c(0.2, 0.6),
        legend.box.background = element_rect(color = 'white'),
        legend.background = element_rect(color = 'grey90'))
        # legend.title = element_blank())
```

```{r}
# plot of demographic data
demographics %>% 
  select(-c('ID', 'survey_weight')) %>%
  mutate_all(as.character) %>% 
  pivot_longer(cols = everything()) %>% 
  ggplot(aes(x = value)) +
  geom_histogram(stat = 'count') +
  facet_wrap(~ name, scales = 'free') +
  labs(title = 'Demographics', 
       x = NULL,
       y = NULL)
```

## Correlations and distances between activities

```{r}
# correlation between categories
atus_long %>% 
  pivot_wider(values_from = value, names_from = description) %>% 
  select(-ID) %>% 
  as.matrix() %>% 
  cor() %>% 
  reshape::melt() %>% 
  ggplot(aes(x = reorder(X1, value), y = reorder(X2, value), fill = value)) +
  geom_tile() +
  scale_fill_gradient2() +
  labs(title = "Correlations between activities",
       x = NULL,
       y = NULL) +
  theme(axis.text.x = element_text(angle = 40, hjust = 1),
        legend.title = element_blank())
```


```{r}
# distance between categories
atus_long %>% 
  pivot_wider(values_from = value, names_from = description) %>% 
  select(-ID) %>% 
  as.matrix() %>% 
  t() %>% 
  dist() %>% 
  as.matrix() %>% 
  reshape::melt() %>% 
  ggplot(aes(x = reorder(X1, -value), y = reorder(X2, -value), fill = value)) +
  geom_tile() +
  scale_fill_gradient2(labels = scales::comma) +
  labs(title = "Distance between activities",
       x = NULL,
       y = NULL) +
  theme(axis.text.x = element_text(angle = 40, hjust = 1),
        legend.title = element_blank())
```


```{r} 
# check for NAs
dim(na.omit(atus_long)) == dim(atus_long)

# look at the data
atus_long %>% 
  ggplot(aes(x = value)) +
  geom_density() +
  facet_wrap(~description, scales = "free") + 
  labs(title = 'Feature densities (un-transformed)', 
       x = NULL,
       y = NULL)
```

## Don't need transformations or scaling because already on the same unit: minutes

# pivot wider and check correlations between activities

```{r}
# pivot wider and turn into matrix
atus_wide <- atus_long %>% 
  pivot_wider(values_from = value, names_from = description) %>% 
  select(-ID) %>% 
  as.matrix()

# distance between categories
atus_wide %>% 
  t() %>% 
  dist() %>% 
  as.matrix() %>% 
  reshape::melt() %>% 
  ggplot(aes(x = reorder(X1, -value), y = reorder(X2, -value), fill = value)) +
  geom_tile() +
  scale_fill_gradient2(labels = scales::comma) +
  labs(title = "Distance between activities",
       x = NULL,
       y = NULL) +
  theme(axis.text.x = element_text(angle = 40, hjust = 1),
        legend.title = element_blank())
```


## PCA 

```{r}
# run PCA
atus_pca <- prcomp(atus_wide)
summary(atus_pca)$importance %>% 
    pander::pander(justify = c('left', rep('right', 15)), round = 2)

pca_plot <- atus_pca$x[ , 1:3]
# rgl::plot3d(pca_plot)
```


## Resampling the data using survey weights

```{r}
# sample using survey weights
total_rows <- nrow(atus_wide)
sample_size <- 25000
rows_to_keep <- sample(1:total_rows, size = sample_size, prob = demographics$survey_weight, replace = TRUE)
IDs_kept <- atus_long %>% 
  pivot_wider(values_from = value, names_from = description) %>% 
  select(ID) %>% 
  .[rows_to_keep,]
atus_resampled <- atus_wide[rows_to_keep,]

# correlations
corr_matrix <- cor(atus_resampled)
corr_matrix[lower.tri(corr_matrix)] %>% density() %>% plot(main = 'Density of correlations between activities')

# run PCA
atus_pca <- prcomp(atus_resampled)
summary(atus_pca)$importance %>% 
      pander::pander(justify = c('left', rep('right', 15)), round = 2)
```

```{r include=FALSE}
# memory management -------------------------------------------------------
rm(corr_matrix, pca_plot, total_rows)
gc()
```

# Clustering
## Hierarchical cluster

```{r}
# distance matrix for features
dist_sc <- dist(atus_resampled, method = 'euclidean')

# ward (D2) linkage hier clustering
hcl_ward <- hclust(d = dist_sc, method = 'ward.D2')
dend <- as.dendrogram(hcl_ward)
hcl_k <- 6
# dend_col <- color_branches(dend, k = hcl_k)
# plot(dend_col, main = paste0('Ward (D2) Linkage: K = ', hcl_k))
groupings <- cutree(hcl_ward, hcl_k)
```

```{r}
# save the cluster memberships
IDs_kept %>% 
  bind_cols(cluster = groupings) %>% 
  write_csv('Analyses/25k_6_cluster_IDs.csv')
```


```{r}
# cleaned up dendrogram
factoextra::fviz_dend(
  cut(dend, h = 2000)$upper,
  k = hcl_k,
  # k_colors = c("#2E9FDF", "#00AFBB", "#E7B800", "#FC4E07"),
  color_labels_by_k = TRUE,
  show_labels = FALSE,
  ggtheme = NULL,
  lwd = 0.5,
) + 
  scale_y_continuous(labels = scales::comma) +
  labs(title = 'Ward (D2) linkage',
       subtitle = paste0(hcl_k, ' cluster solution'),
       x = NULL,
       y = NULL) +
  theme(axis.ticks = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())
```

#### manually building ggplot dendrogram
#### this confirms that dendrogram clusters are not left-to-right 1:6

```{r}
library(ggdendro)

# basic, full tree
dend_data <- dendro_data(dend, type = "rectangle")
ggplot(dend_data$segments) + 
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend))

# tree with labels colored by cluster
dend_data <- dendro_data(dend, type = "rectangle")
groupings <- cutree(hcl_ward, hcl_k)
# join labels with cluster
dend_data$labels <- dend_data$labels %>%
  left_join(data.frame(cluster = as.factor(groupings),
                       label = as.factor(1:length(groupings))))
# plot tree with labels colored
ggplot(dend_data$segments) +
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend)) +
  # geom_text(data = dend_data$labels, aes(x, y, label = label, color = cluster),
  #           hjust = 1, angle = 90, size = 5) +
  geom_tile(data = dend_data$labels, aes(x = x, y = y + 1000, width = 1.1, height = 2000, fill = cluster)) +
  scale_x_continuous(labels = NULL) +
  scale_y_continuous(labels = scales::comma) +
  coord_cartesian(ylim = c(2500, 26000)) +
  labs(title = 'Clipped tree with colors representing clusters at the bottom',
       x = NULL,
       y = NULL) +
  theme(plot.margin = unit(c(1, 1, 1, 1), "cm"))
  

# trimmed, colored branches tree
dend_cut <- cut(dend, h = 2000)$upper
dend_data_trim <- dendro_data(dend_cut, type = "rectangle")

# cluster_membership <- 
#   c(
#     rep(1, 30),
#     rep(NA, 2),
#     rep(2, 66-32),
#     rep(NA, 2),
#     rep(3, 106-68),
#     rep(NA, 2),
#     rep(4, 30),
#     rep(NA, 2),
#     rep(5, 14),
#     rep(6, nrow(dend_data_trim$segments) - 154)
#   )

dend_data_trim$segments$cluster <- as.factor(cluster_membership)
linetypes <- c('solid', 'dashed')[as.numeric(is.na(dend_data_trim$segments$cluster))+1]

ggplot(dend_data_trim$segments) + 
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend,
               color = cluster),
               linetype = linetypes) +
  scale_x_continuous(labels = NULL) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = 'Ward (D2) linkage',
       subtitle = '6 cluster solution',
       x = NULL,
       y = NULL) +
  theme(axis.ticks = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        legend.position = 'none')



## attempt to automate coloring of branches

cluster_max_min_x <- dend_data$labels %>% 
  group_by(cluster) %>% 
  summarize(max_x = max(x),
            min_x = min(x))

return_cluster <- function(x, xend){
  index <- map2_lgl(.x = cluster_max_min_x$min_x, 
                    .y = cluster_max_min_x$max_x, 
                    .f = function(min_x, max_x){
  (x >= min_x & x <= max_x & xend >= min_x & xend <= max_x)
})
  cluster <- cluster_max_min_x$cluster[index]
  if (length(cluster) == 0) cluster <- NA
  
  return(cluster)
}

return_cluster(271.59292,  271.59292)

dend_data_trim$segments <- dend_data_trim$segments %>% 
  rowwise() %>%
  mutate(cluster_auto = return_cluster(x, xend)) %>% 
  ungroup()

# overide non-cluster segments
dend_data_trim$segments$cluster_auto[c(31, 32, 67, 68, 107, 108, 139, 140)] <- NA
# which(is.na(cluster_membership))
# fix a few manually
dend_data_trim$segments$cluster_auto[c(1, 33, 141)] <- c(1, 3, 4)
# !(which(is.na(dend_data_trim$segments$cluster_auto)) %in% which(is.na(cluster_membership)))

ggplot(dend_data_trim$segments) + 
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend,
               color = cluster_auto),
               linetype = linetypes) +
  scale_x_continuous(labels = NULL) +
  scale_y_continuous(labels = scales::comma) +
  scale_color_discrete(name = 'Cluster') +
  labs(title = 'Ward (D2) linkage with correct cluster labels',
       subtitle = '6 cluster solution',
       x = NULL,
       y = NULL) +
  theme(axis.ticks = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank())
  
ggsave(filename = "Analyses/Plots/dendrogram_colored_by_cluster.png",
       width = 8,
       height = 5)
```

```{r}
library(dendextend)
# library(ggdendro)
# Create a complex dend:
dend <- iris[1:30, -5] %>%
  dist() %>%
  hclust() %>%
  as.dendrogram() %>%
  set("branches_k_color", k = 3) %>%
  set("branches_lwd", c(1.5, 1, 1.5)) %>%
  set("branches_lty", c(1, 1, 3, 1, 1, 2)) %>%
  set("labels_colors") %>%
  set("labels_cex", c(.9, 1.2))
# plot the dend in usual "base" plotting engine:
plot(dend)
# Now let's do it in ggplot2 :)
ggd1 <- as.ggdend(dend)
ggplot(ggd1)
```


```{r}

# trimmed, colored branches tree
dend <-  as.dendrogram(hcl_ward) %>% set("branches_k_color", k = hcl_k) %>% set("labels_colors")
dend_cut <- dend #cut(dend, h = 2000)$upper
# dend_data_trim <- dendro_data(dend_cut, type = "rectangle")

ggd1 <- as.ggdend(dend_cut)

# set dashed line for non-cluster segements
ggd1$segments$linetype <- 'solid'
ggd1$segments$linetype[which(is.na(ggd1$segments$col))] <- 'dashed'

text_labels <- tribble(
  ~label, ~x, ~y,
  'Cluster 1', 600, -2000,
  'Cluster 2', 2200, -2000,
  'Cluster 3', 4200, -2000,
  'Cluster 4', 5400, -2000,
  'Cluster 5', 6400, -2000,
  'Cluster 6', 8000, -2000
)

ggplot(ggd1$segments) + 
  geom_segment(aes(x = x, y = y, xend = xend, yend = yend, color = col),
               linetype = ggd1$segments$linetype, alpha = 0.6) +
  geom_text(data = text_labels, aes(label = label, x = x, y = y)) +
  coord_cartesian(ylim = c(0, 26000), clip = 'off') +
  scale_x_continuous(labels = NULL) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = 'Ward (D2) linkage with manual cluster labels',
       subtitle = '6 cluster solution',
       x = NULL,
       y = NULL) +
  theme(axis.ticks = element_blank(),
        panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        legend.position = 'none',
        plot.margin = margin(t = 10, r = 10, b = 30, l = 10, unit = "pt"))

ggsave(filename = "Analyses/Plots/dendrogram_colored_manually.png",
       width = 8,
       height = 5)
```



```{r}
# plot the clusters in PC space
tmp_plot_data <- atus_pca$x[, c(1, 2)] %>% 
  as_tibble() %>% 
  mutate(Cluster = as.factor(groupings))
tmp_plot_data %>% 
  ggplot(aes(x = PC1, y = PC2, color = Cluster, fill = Cluster)) +
  geom_point(alpha = 0.3) +
  geom_polygon(data = get_cluster_polygons(tmp_plot_data, x = "PC1", y = "PC2", cluster = 'Cluster'),
               alpha = 0.1) +
  labs(title = 'Hierarchical Ward D2 cluster solution in PC space',
       subtitle = 'Three cluster solution',
       x = paste0('PC1: ', summary(atus_pca)$importance[2, 'PC1'] %>% scales::percent(accuracy = 0.1), ' variance explained'),
       y = paste0('PC2: ', summary(atus_pca)$importance[2, 'PC2'] %>% scales::percent(accuracy = 0.1),  ' variance explained'))
rm(tmp_plot_data)
```

### PCA pairs plot

```{r eval=FALSE, include=FALSE}
# pairs plot of each PCA component with points colored by cluster
atus_pca$x[, 1:10] %>% 
  as_tibble() %>%
  pairs(., col = as.factor(groupings), upper.panel = NULL)
```


### Examine activities through the clusters

```{r}
# densities of activities by cluster
atus_resampled %>% 
  as_tibble() %>% 
  bind_cols(group = as.factor(groupings)) %>%
  pivot_longer(cols = -group) %>% 
  ggplot(aes(x = value, group = group, fill = group)) +
  geom_density(alpha = 0.3) +
  scale_fill_brewer(palette = 'Accent') +
  scale_x_log10() +
  facet_wrap(~name, nrow = 3, scales = 'free') +
  labs(title = 'Activities split by cluster', 
       x = NULL,
       y = NULL)

ggsave(filename = "Analyses/Plots/activities_by_cluster.png",
       width = 14,
       height = 8)
```

```{r}
# densities of activities by cluster; non-log scale; outliers trimmed
atus_resampled %>% 
  as_tibble() %>% 
  bind_cols(group = as.factor(groupings)) %>%
  pivot_longer(cols = -group) %>%
  group_by(name) %>% 
  filter(value <= quantile(value, 0.99)) %>% 
  ungroup() %>% 
  ggplot(aes(x = value, group = group, fill = group)) +
  geom_density(alpha = 0.3) +
  scale_fill_brewer(palette = 'Accent') +
  facet_wrap(~name, nrow = 3, scales = 'free') +
  labs(title = 'Activities split by cluster', 
       x = NULL,
       y = NULL)
```


### Examine the demographics through the clusters

```{r}
mean_proportions <- demographics %>% 
  right_join(bind_cols(IDs_kept)) %>%
  select(-c('ID', 'survey_weight')) %>%  
  mutate_all(as.character) %>% 
  pivot_longer(cols = everything()) %>% 
  group_by(name, value) %>% 
  summarize(mean_proportion = n() / sample_size)

demographics %>% 
  right_join(bind_cols(IDs_kept, cluster = as.factor(groupings))) %>% 
  select(-c('ID', 'survey_weight')) %>% 
  mutate_all(as.character) %>% 
  pivot_longer(cols = -cluster) %>% 
  group_by(name, cluster) %>% 
  mutate(cluster_n = n()) %>% 
  group_by(name, value, cluster) %>% 
  summarize(proportion = n() / first(cluster_n)) %>% 
  left_join(mean_proportions) %>% 
  mutate(relative_proportion = proportion - mean_proportion) %>% 
  ungroup() %>% 
  # refactor so order is correct
  mutate(value = factor(value,
                        levels = c(
                          as.character(15:85),
                          as.character(0:2),
                          c(
                            "5000",
                            "7500",
                            "10000",
                            "12500",
                            "15000",
                            "20000",
                            "25000",
                            "30000",
                            "35000",
                            "40000",
                            "50000",
                            "60000",
                            "75000",
                            "1e+05",
                            "150000"
                          ),
                          c("asian", "white", "black", "other"),
                          c(
                            "Did not graduate from HS",
                            "HS",
                            "Some college",
                            "Bachelors",
                            "Masters",
                            "Doctoral"
                          )
                        ))) %>% 
  ggplot(aes(x = value, y = relative_proportion, fill = relative_proportion)) +
  geom_col() +
  scale_fill_gradient2(low = 'red', mid = 'grey20', high = 'green') +
  # facet_wrap(cluster ~ name, scales = 'free', ncol = 8) +
  facet_grid(cluster ~ name, scales = 'free_x') +
  labs(title = 'Demographics split by cluster',
       subtitle = 'Relative proportion of values versus population', 
       x = NULL,
       y = NULL) +
  theme(axis.text.x = element_text(angle = 65, hjust = 1),
        axis.text = element_text(size = 8))

ggsave(filename = "Analyses/Plots/demographics_by_cluster.png",
       width = 30,
       height = 20)


demographics %>% 
  right_join(bind_cols(IDs_kept, group = groupings)) %>%
  select(-c('ID', 'survey_weight')) %>%
  mutate_all(as.character) %>% 
  pivot_longer(cols = -group) %>% 
  ggplot(aes(x = value)) +
  geom_histogram(stat = 'count') +
  facet_grid(group ~ name, scales = 'free') +
  labs(title = 'Demographics split by cluster', 
       x = NULL,
       y = NULL)
```

```{r include=FALSE}
# memory management -------------------------------------------------------
rm(dist_sc, groupings, sample_size)
gc()
```

## Notes on cluster demographics

# Archetypes
- Weekend workers:
  - skews younger, sub 30
  - yes children
  - HS and some college
  - across incomes
  - not married
  - across races
  - skews heavily male,
  -Notes: george to research if differneces in employment in high school 
- Education:
  - across ages
  - skews toward having children
  - slight skew towards some college, bachelors, masters
  - skews toward upper income
  - skews married
  - across races
  - skews toward women
- Social, Leisure:
 - skews older
 - heavily no children
 - heavily HS
 - skews lower income
 - skews not married
 - skews black
 - heavy skew male
- Sports, exercise:
  - heavy skew teenages
  - children
  - some HS
  - skews heavy upper income
  - not married
  - heavy white
  - heavy heavy male
- Sleepers + personal care
 - skews younger, sub 30
 - slight skew towards children
 - heavy heavy skew towards did not graduate from HS
 - has child
 - lower income
 - skews not married
 - black
 - slightly female
- Household activities:
  - middle aged
  - slight skew towards children
  - bachelor's
  - upper income
  - heavily skewed towards married
  - white
  - heavy skew female
  

Qs:
- does children / child under 12 mean child in the household, primary care, or adult children?
- look at secondary activities
- how to map clusters back to dendrogram?

- look at data dictionary
- see if there is any outstanding research on these archetypes
- figure out how to make side-by-side publishable plots for each cluster

### Optimizing hierarchical cluster sizes

```{r}
# get optimal cluster sizes: c(g)
hcl_ch <- NbClust(
  data = atus_resampled,
  max.nc = 10,
  method = 'ward.D2',
  index = 'ch'
)
hcl_ch$All.index %>% plot(type = 'l', x = 2:10, main = 'Hierarchical: C(g) index', xlab = 'n clusters')

# get optimal cluster sizes: silhouette width
hcl_silhouette <- NbClust(
  data = atus_resampled,
  max.nc = 7,
  method = 'ward.D2',
  index = 'silhouette'
)
hcl_silhouette$All.index %>% plot(type = 'l', x = 2:7, main = 'Hierarchical: Silhouette', xlab = 'n clusters')


# plot the clusters in PC space
tmp_plot_data <- atus_pca$x[, 1:2] %>% 
  as_tibble() %>% 
  mutate(Cluster = as.factor(hcl_ch$Best.partition))
tmp_plot_data %>% 
  ggplot(aes(x = PC1, y = PC2, color = Cluster, fill = Cluster)) +
  geom_point(alpha = 0.3) +
  geom_polygon(data = get_cluster_polygons(tmp_plot_data, x = PC1, y = PC2, cluster = 'Cluster'),
               alpha = 0.1) +
  labs(title = 'Hierarchical Ward D2 cluster solution in PC space',
       subtitle = paste0(hcl_ch$Best.nc[[1]], ' cluster solution'))
rm(tmp_plot_data)
```

### Examine the demographics through the clusters

```{r}
demographics %>% 
  right_join(bind_cols(IDs_kept, cluster =as.factor(hcl_ch$Best.partition))) %>%
  select(-c('ID', 'survey_weight')) %>%
  mutate_all(as.character) %>% 
  pivot_longer(cols = -cluster) %>% 
  ggplot(aes(x = value)) +
  geom_histogram(stat = 'count') +
  facet_grid(cluster ~ name, scales = 'free') +
  labs(title = 'Demographics split by cluster', 
       subtitle = 'Hierarchical Ward D2 cluster solution',
       x = NULL,
       y = NULL)
```

## kmeans clustering

```{r}
# get optimal cluster sizes 
km_ch <- NbClust(
  data = atus_resampled,
  max.nc = 20,
  method = 'kmeans',
  index = 'ch'
)
km_ch$All.index %>% plot(type = 'l', x = 2:20, main = 'Kmeans: C(g) index', xlab = 'n clusters')

# get optimal cluster sizes 
km_silhouette <- NbClust(
  data = atus_resampled,
  max.nc = 20,
  method = 'kmeans',
  index = 'silhouette'
)
km_silhouette$All.index %>% plot(type = 'l', x = 2:20, main = 'Kmeans: Silhouette', xlab = 'n clusters')

# run the final kmeans algo with optimal number of clusters
km_eleven <- kmeans(x = atus_resampled,
                  centers = 6,
                  nstart = 100,
                  iter.max = 30,
                  algorithm = 'Hartigan-Wong')

# plot the clusters in PC space
tmp_plot_data <- atus_pca$x[, 1:2] %>% 
  as_tibble() %>% 
  mutate(Cluster = as.factor(km_eleven$cluster))
tmp_plot_data %>% 
  ggplot(aes(x = PC1, y = PC2, color = Cluster, fill = Cluster)) +
  geom_point(alpha = 0.3) +
  geom_polygon(data = get_cluster_polygons(tmp_plot_data, x = 'PC1', y = 'PC2', cluster = 'Cluster'),
               alpha = 0.1) +
  labs(title = 'K-means cluster solution in PC space')
rm(tmp_plot_data)
```

```{r}
atus_resampled %>% 
  as_tibble() %>% 
  bind_cols(group = as.factor(km_eleven$cluster)) %>%
  pivot_longer(cols = -group) %>% 
  ggplot(aes(x = value, group = group, fill = group)) +
  geom_density(alpha = 0.3) +
  scale_x_log10() +
  facet_wrap(~name, nrow = 3, scales = 'free') +
  labs(title = 'Activities split by cluster', 
       x = NULL,
       y = NULL)
```


### Examine the demographics through the clusters

```{r}
demographics %>% 
  right_join(bind_cols(IDs_kept, group = as.factor(km_eleven$cluster))) %>%
  select(-c('ID', 'survey_weight')) %>%
  mutate_all(as.character) %>% 
  pivot_longer(cols = -group) %>% 
  ggplot(aes(x = value)) +
  geom_histogram(stat = 'count') +
  facet_wrap(group ~ name, scales = 'free', ncol = 8) +
  labs(title = 'Demographics split by cluster', 
       x = NULL,
       y = NULL)
```

### Kmeans gap statistic

```{r eval=FALSE, include=FALSE}
gc()
# takes an hour !!!!
gap_stat <- cluster::clusGap(atus_resampled, FUN = kmeans, K.max = 20, nstart = 20, B = 25, iter.max = 20)

factoextra::fviz_gap_stat(gap_stat)
```

